A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-10-14 doc/large_strain.tex
More improvements

% \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
%\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{bm}
%\usepackage{bbm}
%\usepackage{caption}
\usepackage{xcolor}
%\usepackage{geometry}
%\usepackage{graphicx}
%\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
%\usepackage{layouts}
%\usepackage{mathrsfs}
%\usepackage{pgfkeys,pgfmath,pgfcore}
%\usepackage{srcltx}
%\usepackage[labelformat=simple]{subcaption}
%\usepackage{xspace}
\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}

@@

  \begin{abstract}
    The performance of finite element solvers on modern computer architectures is typically memory bound.
    The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
    In order to improve the performance of the iterative solver, so-called matrix-free methods are
    widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly.

    In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
    We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
    In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators
    and employ them to define a geometric multigrid preconditioner.
    Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
  \end{abstract}

@@

The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of the iterative solver, so-called matrix-free methods are
widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}. Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
{\color{red}Daniel: please add more links if you want, maybe GPU related}.

To the best of our knowledge, the matrix-free methods are not yet widely adopted within the solid mechanics community.
We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are
more elaborate as compared to the fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case. Another reason could be the smoothness of the underlying solution. FEM with linear or quadratic elements, that avoid locking, are often adopted in solid mechanics community. Higher order elements are rarely employed. However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore we believe that for large scale computations in solid mechanics it is crucial to adopt matrix-free approaches and develop efficient multigrid solvers.

In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
and employ them to define a geometric multigrid preconditioner.
Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.

The paper is organized as follows. In Section \ref{sec:theory} we briefly introduce partial differential equations of finite-strain solid mechanics. Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the proposed matrix-free operator evaluation approaches. In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity. The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions in Section \ref{sec:example}. Summary is presented in Section \ref{sec:summary}.

@@

\subsection{Weak form}
%
% Kinematics
%
Deformation of the body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in Euclidean space $\mathbb E^3$.
The location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

%
% Deformation gradient
%
Linear deformation map is assumed
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient.
The mapping has to be one-to-one and exclude a self penetration, consequently the Jacobian has to be positive $J = \det \gz F > 0$.
Deformation gradient can be split in the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors as $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
\begin{align}
  \chi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}
\end{align}

%
% strain measures
%
In what follows we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$. Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$. We will also use the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T = \gz V^2$.

For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

The principal of stationary potential energy requires that at the equilibriums the directional derivative with respect to the displacement $\gz u$
\begin{align}
\delta \mcl E =
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the first Piola-Kirchhoff stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
(\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
\gz \tau : (\Grad \, \delta \gz u \cdot \gz F^{-1}) =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to apply the Newton method, linearization of \eqref{eq:weak_form} around a given solution field $\overline{\gz u}$ is required. This is given by
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V
\end{align}
This can be simplified \cite{Wriggers2008} to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$ and the fourth order tangent tensor is push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is called material part of the tangent operator as it is related to the linearization of the Kirchhoff stress. Whereas the second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from linearization of the deformation gradient $\gz F$.

@@

For the current study we use the Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ respectively denote the shear modulus and Lam\'{e} parameter. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated material tangent are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}
\label{eq:simplified_action}
\end{gather}

@@

We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in a
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
For the sake of this study $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.

>>>>

% \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
%\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{bm}
%\usepackage{bbm}
%\usepackage{caption}
\usepackage{xcolor}
%\usepackage{geometry}
%\usepackage{graphicx}
%\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it is 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
%\usepackage{layouts}
%\usepackage{mathrsfs}
%\usepackage{pgfkeys,pgfmath,pgfcore}
%\usepackage{srcltx}
%\usepackage[labelformat=simple]{subcaption}
%\usepackage{xspace}
\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}

@@

  \begin{abstract}
    The performance of finite element solvers on modern computer architectures is typically memory bound.
    The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
    In order to improve the performance of the iterative solver, so-called matrix-free methods are
    widely adopted in the fluid mechanics community, where the matrix-vector products are computed on-the-fly.

    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyper-elastic tangent operator numerically.
    In order to improve the convergence behavior of iterative solvers, we also propose a way to construct level tangent operators
    and employ them to define a geometric multigrid preconditioner.
    Our implementation employs MPI {\color{red}and Intel Threading Building Blocks parallelization} and SIMD vectorization.
    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
  \end{abstract}

@@

The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of the iterative solver, so-called matrix-free methods are
widely adopted in fluid mechanics community, where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}. Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
A possible reason is that the tangent operators (coming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used in fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case. Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution. FEM with linear or quadratic elements, that avoid locking, are often adopted in the solid mechanics community. Higher order elements are rarely employed. However, we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

In this work, we investigate different implementations of the finite-strain hyper-elastic tangent operator with respcet to their computational cost.
In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
and employ them to define a geometric multigrid preconditioner.
Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.

The paper is organized as follows. In Section \ref{sec:theory}, we briefly introduce the partial differential equations used in  finite-strain solid mechanics. Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches. In Section \ref{sec:gmg}, we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity. The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions in Section \ref{sec:example}. The results are summarized in Section \ref{sec:summary}.

@@

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in the Euclidean space $\mathbb E^3$.
The location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

%
% Deformation gradient
%
We assume that the deformation map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian has to be positive $J = \det \gz F > 0$. The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
\begin{align}
  \chi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}

%
% strain measures
%
In what follows, we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$. Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$. We will also use the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T = \gz V^2$.

For conservative systems without volumetric load, the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

The principal of stationary potential energy requires for all  equilibrium states $\gz u$ that the directional derivative with respect to the displacement
\begin{align}
\delta \mcl E =
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the first Piola-Kirchhoff stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
(\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
\gz \tau : (\Grad \, \delta \gz u \cdot \gz F^{-1}) =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
\end{align}
Following \cite{Wriggers2008}, this can be simplified to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the the linearization of the deformation gradient $\gz F$.

@@

For the current study, we use the Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated material tangent are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}

@@

We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in a
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}

The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to obtained as solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.
